Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-04-14 endofquarterpaper.qmd
merge

The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020], coupled with a biological transition from Neanderthals to early modern humans. Archaeologists have frequently relied upon variations in artwork, the large rise of which is a distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile objects from the Upper Paleolithic in Europe across four phases of the Aurignacian period.

@@

time_unique_data_full$time_period <- factor(time_unique_data_full$time_period, 
                                            levels = c("transitional", "proto_aurignacian", 
                                                       "early_aurignacian", "evolved_aurignacian"), 
                                            labels = c("Transitional", "Proto-Aurignacian",
                                                       "Early Aurignacian", "Evolved Aurignacian"))
```

```{r}
trans_site_count <- nrow(trans_unique_data)
trans_sign_count <- ncol(trans_artifact_data)
proto_site_count <- nrow(proto_unique_data)
proto_sign_count <- ncol(proto_artifact_data)
early_site_count <- nrow(early_unique_data)
early_sign_count <- ncol(early_artifact_data)
evolved_site_count <- nrow(evolved_unique_data)
evolved_sign_count <- ncol(evolved_artifact_data)

@@

sign_material_df <- signbase_full_clean %>% 
  group_by(material) %>% 
  summarise(count = n())%>% 
  arrange(desc(count))

iv_count <- sign_material_df %>% 
  filter(material == "ivory") %>% 
  pull(count)

bone_count <-sign_material_df %>% 
  filter(material == "bone") %>% 
  pull(count)

lime_count <- sign_material_df %>% 
  filter(material == "limestone") %>% 
  pull(count)

ant_count <- sign_material_df %>% 
  filter(material == "antler") %>% 
  pull(count)

tooth_count <- sign_material_df %>% 
  filter(material == "animal tooth") %>% 
  pull(count)

ochre_count <- sign_material_df %>% 
  filter(material == "ochre") %>% 
  pull(count)

undet_count <- sign_material_df %>% 
  filter(material == "undetermined") %>% 
  pull(count)

stone_row <- tibble("material" = c("stone"), "count" = c(1))

sign_material_df <- sign_material_df %>% 
  rbind(stone_row) %>% 
  filter(material != "stone indet.")

misc_string <- sign_material_df %>% 
  filter(count == 1) %>% 
  dplyr::select(material) %>% 
  c()

misc_string <- misc_string$material

object_type_df <- signbase_full_clean %>% 
  group_by(object_type) %>% 
  summarise(count = n()) %>% 
  arrange(desc(count))
```

@@

## mantel chronological
chron_df <- signbase_full_clean %>% 
  column_to_rownames("id") %>% 
  dplyr::select(MedianBP) %>% 
  dist(method = "euclidean")

chron_sign_df <- signbase_full_clean %>% 
  column_to_rownames("id") %>% 
  dplyr::select(line:star) %>% 
  mutate_all(~ as.numeric(. > 0)) %>% 
  vegdist("jaccard")

@@

Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe. SignBase 1.0 is a record of objects noting only the presence or absence of each sign on the object, without details of the quantity or sequence of individual signs on a given object. We used the identification of sign types provided by SignBase's curators [@dutkiewicz2020]. The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types. We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities. We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II). These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map). `r iv_count` of the objects are made of ivory, `r bone_count` objects are made of bone, `r lime_count` objects are made of limestone, `r ant_count` objects are made of antler, `r tooth_count` objects are made of animal teeth, `r ochre_count` objects are made of ochre, and `r undet_count` objects are made of undetermined material. The rest of the objects are made of `r combine_words(misc_string)`, with one made of each material.

@@

The distribution of objects across sites is highly skewed, as shown in @fig-site-map C. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types in order to make sure that the results would not be significantly impacted by the exlcusion of object abundance. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The Mantel tests, performed using the vegan package [@Vegan2001], indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types. A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`), confirming that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.

@@

If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases. This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geographic distance for each phase using Mantel tests. The results as shown in @tbl-mantel-table show that none of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.

@@

The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are minimal. The Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and the Evolved Aurignacian has the weakest groupings, with a score of `r evolved_modularity`. There is an increase in modularity between the Transitional period and the Proto-Aurignacian, before an decrease into the Early Aurignacian, followed by an increase with the Evolved Aurignacian.

@@

The results also add to our knowledge of social networks within the Aurignacian period, suggesting dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 place the emergence of inter-group social networks at around 30 k BP, however, our network analysis shows evidence of inter-group social networks beginning at around 41 k BP, suggesting an earlier emergence than previously predicted. Dividing the data into smaller time periods potentially serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023; @Daems_Coco_Gillreath-Brown_Kafetzaki_2024]. Additionally, by focusing on change over time, we can see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019]

>>>>

The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution. Another distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile objects from the Upper Paleolithic in Europe across four phases of the Aurignacian period.

@@

time_unique_data_full$time_period <- factor(time_unique_data_full$time_period, 
                                            levels = c("transitional", "proto_aurignacian", 
                                                       "early_aurignacian", "evolved_aurignacian"), 
                                            labels = c("Transitional", "Proto-Aurignacian",
                                                       "Early Aurignacian", "Evolved Aurignacian"))

trans_site_count <- nrow(trans_unique_data)
trans_sign_count <- ncol(trans_artifact_data)
proto_site_count <- nrow(proto_unique_data)
proto_sign_count <- ncol(proto_artifact_data)
early_site_count <- nrow(early_unique_data)
early_sign_count <- ncol(early_artifact_data)
evolved_site_count <- nrow(evolved_unique_data)
evolved_sign_count <- ncol(evolved_artifact_data)

nvogel <- nrow(signbase_full_clean %>% 
                 filter(site_name == "Vogelherd"))

@@

artifact_count_median <- median(artifact_count_df$artifact_count)

@@

## mantel chronological
chron_df <- signbase_full_clean %>% 
  column_to_rownames("id") %>% 
  dplyr::select(MedianBP) %>% 
  dist( method = "euclidean")

chron_sign_df <- signbase_full_clean %>% 
  column_to_rownames("id") %>% 
  dplyr::select(line:star) %>% 
  # BM: Ensure binary data as input to jaccard distance
  mutate_all(~ as.numeric(. > 0)) %>% 
  vegdist("jaccard")

@@

Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe. SignBase 1.0 is a record of objects noting only the presence or absence of each sign on the object, without details of the quantity or sequence of individual signs on a given object. We used the identification of sign types provided by SignBase's curators [@dutkiewicz2020]. The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types. We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities. We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II). These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).

@@

The distribution of objects across sites is highly skewed, as seen in @fig-site-map C. To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites. We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types. The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value. The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation. The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The Mantel tests were performed using the vegan package [@Vegan2001]. The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).

Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table. The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types. A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`). This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.

@@

If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases. This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase. We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table. None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.

@@

The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are minimal. The Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and the Evolved Aurignacian has the weakest groupings, with a score of `r evolved_modularity`. There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger increase with the Evolved Aurignacian.

@@

This study adds to our knowledge of social networks within the Aurignacian period, as our results suggest dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our network analysis shows evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted. We hope that dividing the data up into smaller time periods serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023]. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].
